Supplementary Information, Possession trance covaries with measures of social rigidity in the Ethnographic Atlas

### 2023-05-14 Readme.Rmd
new final draft 2

This is the online supplement for the paper “Possession trance covaries
with measures of social rigidity in the Ethnographic Atlas”. The
current version of the paper is
[here](https://docs.google.com/document/d/1qbcsCeQ-Tw2W-yhQvvYd1Nk5ImQznqWBmiHX09lB_24/edit?usp=sharing). The `helper` file creates the data files. The `model` files run the Gradient Boosting Model and the Bayestraits model, respectively.

@@

Here is a map of trance and possession phenomena.

```{r fig.width=8, fig.height=11}
lld = ld %>% 
  filter(var_id == 'EA112') %>% 
  mutate(
    lon2 = ifelse(lon < -20, lon + 360, lon),
    coding = fct_reorder(var_name,code),
    var_description2 = glue('({code}) {var_description}')
  )

@@

```{r }
trance = sccs %>% 
  filter(var_id == 'EA112') %>% 
  count(var_name,code,var_description) %>% 
  rename(name = var_name, description = var_description, "n in SCCS" = n) %>% 
  mutate(
    possession_trance = case_when(
      code %in% c(3,4,6,7) ~ 'possession trance',
      code %in% c(1,2,8,5) ~ 'other'
    ) %>% 
      fct_relevel('possession trance')
  )

@@

```{r }
trance %>% 
  group_by(possession_trance) %>% 
  summarise(count = sum(`n in SCCS`)) %>% 
  kable('simple')
```

@@

```{r eval = F}
pred_table = ld %>% 
  filter(var_title != 'Trance states') %>% 
  distinct(var_id,var_title,var_definition) %>%
  arrange(var_id) %>% 
  rename(ID = var_id, Name = var_title, Definition = var_definition)

@@

```{r }
pred_table = read_tsv('data/pred_table_added.tsv')

@@

```{r }
cord = ld %>% 
  filter(!var_title %in% c('Trance states','Societal rigidity')) %>% 
  select(soc_id,var_title,code) %>%
  pivot_wider(id_cols = soc_id, names_from = var_title, values_from = code) %>% 
  select(-soc_id)

@@

The predictors are the possible covariates identified above. Some predictors, like `EA032` (Jurisdictional hierarchy of local community) or `EA113` can be construed to follow a scale. These are input as *ordered factors*. Others, like `EA053` (Sex differences in animal husbandry) or `EA043` (Major type of descent) can be carved up into categories: These are *one-hot encoded*. For details, see the Appendix.

@@

```{r }

@@

```{r }
chisq.test(c(c(68,12),c(7,64))) %>% 
  tidy() %>% 
  kable('simple', digits = 2)
```

@@

```{r }

@@

```{r }
chisq.test(c(c(264,113),c(87,194))) %>% 
  tidy() %>% 
  kable('simple', digits = 2)
```

@@

```{r }
vars = ld %>% 
  filter(var_title != 'Trance states') %>% 
  distinct(var_id,var_title,var_definition,var_name) %>%
  arrange(var_id) %>% 
  group_by(var_id, var_title, var_definition) %>% 
  nest() %>% 
  mutate(levels = map(data, ~ paste(.$var_name, collapse = ', '))) %>% 
  select(var_id, var_title, levels) %>% 
  unnest(cols = c(levels)) %>% 
  rename(ID = var_id, Name = var_title)

@@

```{r fig.width=10,fig.height=16}
d_out = ld %>% 
  filter(var_id == 'EA112') %>% 
  mutate(
    `possession trance` = case_when(
      code %in% c(3,4,6,7) ~ 'present',
      code %in% c(1,2,5,8) ~ 'absent'
        )
      ) %>% 
  select(soc_id,`possession trance`)

@@

@@

In the resulting fit, we see that six of the seven predictors have a robust effect on the outcome, with a 95% credible interval that excludes zero. All of the predictors are ordered categories and sometimes the effect is linear, sometimes it's quadratic, or cubic. The exception is premarital sexual norms for girls (EA078).

@@

## 7. Did these traits develop independently?

@@

We select two phylogenies, Austronesian and Bantu. We use BayesTraits to assess the co-evolution of predictors and outcome (@pagel2007bayestraits). We fit two models, an independent model, which assumes that two traits evolve independently, and a dependent model, which assumes that traits are correlated and the rate of change in one trait. is dependent on the state of the other trait.

@@

bayes factor
```{r }
#write_sheet(bf, ss = 'https://docs.google.com/spreadsheets/d/1X7ZCnw6LVhjLIy4SiSeK3iZ6o1S144iHMv4DqOaCDEU/edit#gid=0', sheet = 'Table 7')
kable(bf, 'simple', digits = 2)
```

@@

## 8. Summary

@@

## 9. Limitations

The cross-cultural model is fit on the Standard Cross-Cultural Sample which does not minimise the phylogenetic signal completely (see [here](https://excd.org/2018/03/01/stats-corner-is-the-standard-cross-cultural-sample-really-standard/)). The data in the Ethnographic Atlas cannot be independently verified. The outcome variable and the predictors are selected and defined with the existing literature in mind. It remains true that these decisions use up a large amount of researcher degrees of freedom. This is especially true for the phylogenetic comparative analysis, which further reduces resolution on the co-varying traits as a starting point.

@@

```{r }
kable(vars, 'simple')
```

@@

```{r }
coding_table = ld %>% 
  filter(var_id != 'EA112') %>% 
  distinct(var_id,var_title,var_description,recoded_value) %>% 
  arrange(var_id,recoded_value)

@@

```{r, echo=T}
gbm_params1 = list(
  learn_rate = c(0.01,0.02,0.03,0.1),
  max_depth = c(1,3,5,7),
  sample_rate = c(0.4,0.7,0.9,0.95),
  col_sample_rate = c(0.1,0.2,0.3),
  min_split_improvement = c(1e-3,1e-5)
)

>>>>

This is the online supplement for the paper “Possession trance covaries
with measures of social rigidity in the Ethnographic Atlas”. The
current version of the paper is
[here](https://docs.google.com/document/d/1qbcsCeQ-Tw2W-yhQvvYd1Nk5ImQznqWBmiHX09lB_24/edit?usp=sharing). The `helper` file creates the data files. The `model` files run the Gradient Boosting Model, the Generalised Linear Model, and the Bayestraits model, respectively.

@@

Here is a map of trance and possession phenomena in the Atlas.

```{r possession_trance_map, fig.width=8, fig.height=11}
lld = ld %>% 
  filter(var_id == 'EA112') %>% 
  mutate(
    lon2 = ifelse(lon < -20, lon + 360, lon),
    coding = fct_reorder(var_name,code),
    var_description2 = glue('({code}) {var_description}')
  )

@@

```{r table_1}
trance = sccs %>% 
  filter(var_id == 'EA112') %>% 
  count(var_name,code,var_description) %>% 
  rename(name = var_name, description = var_description, "n in SCCS" = n) %>% 
  mutate(
    possession_trance = case_when(
      code %in% c(3,4,6,7) ~ 'possession trance',
      code %in% c(1,2,8,5) ~ 'other'
    ) %>% 
      fct_relevel('possession trance')
  )

@@

```{r table_2}
trance %>% 
  group_by(possession_trance) %>% 
  summarise(count = sum(`n in SCCS`)) %>% 
  kable('simple')
```

@@

```{r table_3, eval = F}
pred_table = ld %>% 
  filter(var_title != 'Trance states') %>% 
  distinct(var_id,var_title,var_definition) %>%
  arrange(var_id) %>% 
  rename(ID = var_id, Name = var_title, Definition = var_definition)

@@

```{r table_3_ctd}
pred_table = read_tsv('data/pred_table_added.tsv')

@@

```{r corrplot}
cord = ld %>% 
  filter(!var_title %in% c('Trance states','Societal rigidity')) %>% 
  select(soc_id,var_title,code) %>%
  pivot_wider(id_cols = soc_id, names_from = var_title, values_from = code) %>% 
  select(-soc_id)

@@

The predictors are the possible covariates identified above. Some predictors, like `EA032` (Jurisdictional hierarchy of local community) or `EA113` can be construed to follow a scale. These are input as *ordered factors*. Others, like `EA053` (Sex differences in animal husbandry) or `EA043` (Major type of descent) can be carved up into categories: These are *one-hot encoded*. So, for `EA043`, this means that the category originally has four levels, such as “patrilineal”, “matrilineal”, “cognatic”, and “other”, and it will be converted into three categories, which are `true` or `false` for patrilineal, matrilineal, and cognatic, respectively. If all three categories are false, the category has the fourth level, which i “other”. For details, see the Appendix.

@@

```{r table_4}

@@

```{r table_4_ctd}
chisq.test(c(c(68,12),c(7,64))) %>% 
  tidy() %>% 
  kable('simple', digits = 2)
```

@@

```{r table_5}

@@

```{r table_5_ctd}
chisq.test(c(c(264,113),c(87,194))) %>% 
  tidy() %>% 
  kable('simple', digits = 2)
```

@@

```{r table_6}
vars = ld %>% 
  filter(var_title != 'Trance states') %>% 
  distinct(var_id,var_title,var_definition,var_name) %>%
  arrange(var_id) %>% 
  group_by(var_id, var_title, var_definition) %>% 
  nest() %>% 
  mutate(levels = map(data, ~ paste(.$var_name, collapse = ', '))) %>% 
  select(var_id, var_title, levels) %>% 
  unnest(cols = c(levels)) %>% 
  rename(ID = var_id, Name = var_title)

@@

```{r bigplot, fig.width=10,fig.height=16}
d_out = ld %>% 
  filter(var_id == 'EA112') %>% 
  mutate(
    `possession trance` = case_when(
      code %in% c(3,4,6,7) ~ 'present',
      code %in% c(1,2,5,8) ~ 'absent'
        )
      ) %>% 
  select(soc_id,`possession trance`)

@@

### 7. Validating the EA results

@@

In the resulting fit, we see that six of the seven predictors have a robust effect on the outcome, with a 95% credible interval that excludes zero. All of the predictors are ordered categories and sometimes the effect is linear, sometimes it's quadratic, or cubic. There is one predictor that was deemed important by the boosting model but doesn't do a lot here. This is premarital sexual norms for girls (EA078).

@@

If anything, the results are promising. Using one set of data and one approach, we identified a set of relevant predictors. Using a different (larger) set of data and another approach, we confirmed that most of these seem to be useful.

### 8. Did these traits develop independently?

@@

We select two phylogenies, Austronesian (@gray_language_2009) and Bantu (@grollemund_bantu_2015). We use BayesTraits to assess the co-evolution of predictors and outcome (@pagel2007bayestraits). We fit two models, an independent model, which assumes that two traits evolve independently, and a dependent model, which assumes that traits are correlated and the rate of change in one trait. is dependent on the state of the other trait.

@@

```{r table_7}
#write_sheet(bf, ss = 'https://docs.google.com/spreadsheets/d/1X7ZCnw6LVhjLIy4SiSeK3iZ6o1S144iHMv4DqOaCDEU/edit#gid=0', sheet = 'Table 7')
kable(bf, 'simple', digits = 2)
```

@@

### 9. Summary

@@

## 10. Limitations

The cross-cultural model is fit on the Standard Cross-Cultural Sample which does not minimise the phylogenetic signal completely (@eff_does_2004). The data in the Ethnographic Atlas cannot be independently verified, though it has some external validity (@bahrami-rad_tabulated_2021). Using a hierarchical model, as we did here, goes some way in reducing the phylogenetic signal, but this could be improved in the longer term (@claessens_non-independence_2022). The outcome variable and the predictors are selected and defined with the existing literature in mind. It remains true that these decisions use up a large amount of researcher degrees of freedom. This is especially true for the phylogenetic comparative analysis, which further reduces resolution on the co-varying traits as a starting point.

@@

```{r app_table_vars}
kable(vars, 'simple')
```

@@

```{r app_table_coding}
coding_table = ld %>% 
  filter(var_id != 'EA112') %>% 
  distinct(var_id,var_title,var_description,recoded_value) %>% 
  arrange(var_id,recoded_value)

@@

```{r app_table_hyp, echo=T}
gbm_params1 = list(
  learn_rate = c(0.01,0.02,0.03,0.1),
  max_depth = c(1,3,5,7),
  sample_rate = c(0.4,0.7,0.9,0.95),
  col_sample_rate = c(0.1,0.2,0.3),
  min_split_improvement = c(1e-3,1e-5)
)
